Tests of measurement invariance via stochastic processes: {S}ome properties

### 2013-08-14 frontiers/frontiers.Rnw
update refs.bib for reference. go through the connection part in tutorials.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@238 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

where $x_{i}$ is the data vector for individual $i$, $t_{i}$ is the latent variable 
for variable $i$ that the scales purport to measure, and $f$ is the model's 
distributional form {addref}. If the cut point of $v$ is known in advance, i.e. 
gender, nested multiple group models {addref} coupled with likelihood ratio test 
are most commonly applied. Suppose there are two groups A and B. The null hypothesis 
of measurement invariance is

@@

has an asymptotic $\chi^{2}$ distribution with degrees of freedom equal 
to the number difference of parameters in $\theta$.

To be more specific, in the following section, we will introduce how to implement 
multi group analysis in lavaan package. For more details, see {addref}
\subsection{Multiple Group Analysis}
Froh et al had a large sample of youth(n=1401, ranging from late childhood(10 years old) 
to late adolescent(19 years old) complete the three most widely used adult gratitude inventories, 
including Gratitude Questionnaire 6(GQ-6; addref), Gratitude Adjective Checklist(GAC; addref) 
and Gratitude, Resentment, Appreciation Test-Short Form (GRAT-Short Form); addref). 
In this specific analysis, the authors were interested in whether the youth factor 
structure for GQ-6 is tau equivalent model or a congeneric model.

To conduct the analysis, first of all, we need to read in data to R and clean the data.
<<read data>>=
# Reading in the data
data("YouthGratitude",package="psychotools")
# remove cases with 'imputed' values(not in 1,...,9)
yg<-YouthGratitude[apply(YouthGratitude[,4:28],1,function(x)all(x %in% 1:9)),]
@
Second, we want to test which parameter(s) violate measurement invariance. 
In lavaan package, loading, manifest variables' intercept and factor means 
could be tested together or specifically. For more details, 
please see {addref}. Since in most cases researchers are 
interested in the factor loading invariance, we will use that as the 
example in this tutorial.
To implement LRT as ~\ref{equ:LRT}, we need to get estimation 
for $\hat{\theta}$, without differing groups.

<<theta estimate>>=
# estimate theta 
theta<-cfa(
  'f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
  data=yg,group="agegroup",meanstructure=TRUE,
  group.equal="loadings")
@
Then we can get estimate for group A and group B. 
<<theta A and theta B estimate>>=
# estimate theta A,B 
thetaAB<-cfa(
  'f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
  data=yg,group="agegroup",meanstructure=TRUE)
@
In the end, we can get the LRT results by conducting 
log likelihood test, which is done by anova function in R. 
<<LRT>>=
# LRT
anova(theta, thetaAB)
@

\subsection{Extensions for Unknown Group}
The multiple group test performs typically well when the $V$ 
is a relatively-small number of categories and the sample size 
is not too large (LRT is sensitive to large sample and will 
return "false" significance results, details see next section). 
However, when $V$ (in this case, age) continuous variable, however, 
multiple-group analysis usually cannot be performed because there are 
no existing groups. Instead, we can follow the procedure described in
the Background section.  That is, we can fit a model whose parameters 
are restricted to be equal across across all individuals and then examine 
how individuals' scores $s(\hat{theta};x_{i})$ fluctuate with their
values of $V$.

@@

\SweaveOpts{engine = R, eps = TRUE, echo = TRUE, pdf=TRUE}
Alternatively, we could use sctest.lavaan function to get the results.

@@

<<conti3resmaxLM>>=
maxLM<-sctest(theta, as.numeric(yg$agegroup), parm=1:4, stat="maxLM", fplot=FALSE) 
maxLM
@

\subsection{Ordinal Test}
However, in this dataset, $V$ is actually ordinal instead of 
continuous. There is only a partial ordering of individuals 
with respect to $V$,i.e. observations with the same level 
of $V$ have no unique ordering. The ordinal statistics 
proposed here are similar to those described in Equation 
and above, except that we focus on "bins" of individuals 
at each level of the ordinal variable. That is, instead 
of aggregating over all $i=1,...n$ individuals, we 
first compute cumulative proportions $t_{l}(l=1,...,m-1)$ 
associated with the first $m-1$ levels of $V$. 
%% TODO Move these details to Background section
We then aggregate the cumulative scores only over 
$i_{l}=\lfloor n \cdot t_{l} \rfloor$. Test statistics 
related can be written as 
\begin{eqnarray}
    \label{eq:wdm}
      \mathit{WDM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} ~ 
      \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) 
      \right\}^{-1/2}\max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} |,\\
    \label{eq:maxlmo}
      \max \mathit{LM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} 
      ~\left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) 
      \right\}^{-1}\sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2,
\end{eqnarray}

The difference between the continuous test and ordinal 
test is that we don't force arbitary order on the observations 
within the same age groups. Therefore, this ordinal test should 
reflect more of the data. In terms of code, we still need the gefp 
function to get the B estimate. After that, we can use different 
functions to aggregate and scale B estimation. 
The function are called ordwmax and ol2bb1.fun, respectively. 
<<Ordinal Test>>=
gefp1<-gefp(theta,fit=NULL,order.by=as.numeric(yg$agegroup),
       vcov=info_full,sandwich=FALSE,parm=1:4)
pval.ord=rep(NA,2)
pval.ord[1]<-sctest(gefp1,functional=ordwmax(gefp1))$p.value

@@

\begin{figure}
\caption{Get Ordinal Estimate WDMo Alternative Way}
\label{fig:ord1res}
\setkeys{Gin}{width=\textwidth}
<<ord1res, fig=TRUE, height=7, width=7>>=
WDMo<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, stat="WDMo", fplot=TRUE)
WDMo 
@
\end{figure}

@@

In employing the ordinal tests, we obtain $p=0.060$ and $p=0.096$. 
Both $p-$values are clearly larger than that of the 
likelihood ratio test and neither is significant at $\alpha=0.05$

>>>>

where $x_{i}$ is the data vector for individual $i$, $t_{i}$ is the latent variable 
for variable $i$ that the scales purport to measure, and $f$ is the model's 
distributional form \cite{mel89}. If the cut point of $v$ is known in advance, i.e. 
gender, nested multiple group models \cite{bollen98} coupled with likelihood ratio test 
are most commonly applied. Suppose there are two groups A and B. The null hypothesis 
of measurement invariance is

@@

,which has an asymptotic $\chi^{2}$ distribution with degrees of freedom equal 
to the number difference of parameters in $\theta$.

To be more specific, in the following section, we will illustrate with an example to demonstrate how to implement multi group analysis in lavaan package. For more details, see \cite{vande12}

\subsection{Multiple Group Analysis}
Froh et al\citeA{FroFan11} had a large sample of youth(n=1401, ranging from late childhood(10 years old) 
to late adolescent(19 years old) complete the three most widely used adult gratitude inventories, 
including Gratitude Questionnaire 6(GQ-6; \citeNP{MccEmm02}), Gratitude Adjective Checklist(GAC; \citeNP{MccEmm02})  and Gratitude, Resentment, Appreciation Test-Short Form (GRAT-Short Form); \citeNP{ThoWat03}). In this specific analysis, the authors were interested in whether the youth factor 
structure for GQ-6 is a tau equivalent model (factor loadings restricted to be equal across
levels of age)or a congeneric model(all parameters free for each level of age).

To conduct the analysis, first of all, we need to read in data to R and clean the data.
<<read data>>=
# Reading in the data
data("YouthGratitude",package="psychotools")
# remove cases with 'imputed' values(not in 1,...,9)
yg<-YouthGratitude[apply(YouthGratitude[,4:28],1,function(x)all(x %in% 1:9)),]
@
Second, we want to test which parameter(s) violate measurement invariance. 
In lavaan package, loading, manifest variables' intercept and factor means 
could be tested together or specifically. For more details, 
please see \cite{vande12}. Since in most cases researchers are 
interested in the factor loading invariance, we will use that as the 
example in this tutorial.
To implement LRT as ~\ref{equ:LRT}, we need to get estimation 
for $\hat{\theta}$, without differing groups.

<<theta estimate>>=
# estimate theta 
theta<-cfa(
  'f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
  data=yg,group="agegroup",meanstructure=TRUE,
  group.equal="loadings")
@
Then we can get estimate for group A and group B. 
<<theta A and theta B estimate>>=
# estimate theta A,B 
thetaAB<-cfa(
  'f1=~gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5',
  data=yg,group="agegroup",meanstructure=TRUE)
@
In the end, we can get the LRT results by conducting 
log likelihood test, which is done by anova function in R. 
<<LRT>>=
# LRT
anova(theta, thetaAB)
@
Likelihood test is sensitive to large observations (CITE), it is not surprising to 
have a significant result, although Froh et al\citeA{FroFan11} showed that tau equivalent
model has as good alternative fit indices as congeneric model.

\subsection{Extensions for Unknown Group}
The multiple group test performs typically well when the $V$ 
is a relatively-small number of categories and the sample size 
is not too large. However, when $V$ (in this case, age) continuous variable, however, 
multiple-group analysis usually cannot be performed because there are 
no existing groups. Instead, we can follow the procedure described in
the Background section.  That is, we can fit a model whose parameters 
are restricted to be equal across across all individuals and then examine 
how individuals' scores $s(\hat{theta};x_{i})$ fluctuate with their
values of $V$.

@@

\SweaveOpts{engine = R, eps = TRUE, echo = TRUE, pdf=TRUE}
Alternatively, we could use sctest.lavaan function to get the statistics value,
corresponding $p$ value and the plot ordered by age level.

@@

<<conti3resmaxLM>>=
maxLM<-sctest(theta, as.numeric(yg$agegroup), parm=1:4, stat="maxLM", fplot=FALSE) 
maxLM
@
Since there are more than one observations located in each age level. 
So the observation order within each age level is not unique, thus, 
maxLM can't produce the plot. Actually, all three continuous statistics 
were calculated on the arbitrary order that particular dataset entered 
in each age level. In order to accurately account for the multiple observations 
at the same age level, it is more appropriate to use the proposed ordinal test.

\subsection{Ordinal Test}
As a matter of fact, there is only a partial ordering of individuals 
with respect to $V$,i.e. observations with the same level 
of $V$ have no unique ordering. The ordinal statistics 
proposed here are similar to those described in Equation 
and above, except that we focus on "bins" of individuals 
at each level of the ordinal variable. That is, instead 
of aggregating over all $i=1,...n$ individuals, we 
first compute cumulative proportions $t_{l}(l=1,...,m-1)$ 
associated with the first $m-1$ levels of $V$. 
%% TODO Move these details to Background section
We then aggregate the cumulative scores only over 
$i_{l}=\lfloor n \cdot t_{l} \rfloor$. Test statistics 
related can be written as 
\begin{eqnarray}
    \label{eq:wdm}
      \mathit{WDM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} ~ 
      \left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) 
      \right\}^{-1/2}\max_{j = 1, \dots, k} | {\bm B}(\hat {\bm \theta})_{ij} |,\\
    \label{eq:maxlmo}
      \max \mathit{LM}_o & = & \max_{i \in \{i_1, \dots, i_{m-1} \}} 
      ~\left\{ \frac{i}{n} \left( 1 - \frac{i}{n} \right) 
      \right\}^{-1}\sum_{j = 1, \dots, k} {\bm B}(\hat {\bm \theta})_{ij}^2,
\end{eqnarray}

The difference between the continuous test and ordinal 
test is that we don't force arbitrary order on the observations 
within the same age groups. Therefore, this ordinal test should 
reflect more real status of the data. In terms of code, we still need the gefp 
function to get the B estimate. After that, we can use different 
functions to aggregate and scale B estimation. 
The function are called ordwmax and ol2bb1.fun, respectively. 
<<Ordinal Test>>=
gefp1<-gefp(theta,fit=NULL,order.by=as.numeric(yg$agegroup),
       vcov=info_full,sandwich=FALSE,parm=1:4)
pval.ord=rep(NA,2)
pval.ord[1]<-sctest(gefp1,functional=ordwmax(gefp1))$p.value

@@

Or alternatively, we could use the sctest.lavaan function, but assign ordinal statistics.
\begin{figure}
\caption{Get Ordinal Estimate WDMo Alternative Way}
\label{fig:ord1res}
\setkeys{Gin}{width=\textwidth}
<<ord1res, fig=TRUE, height=7, width=7>>=
WDMo<-sctest(theta, as.numeric(yg$agegroup),parm=1:4, stat="WDMo", fplot=TRUE)
WDMo 
@
\end{figure}

@@

In employing the ordinal tests, we obtain $p=0.060$ and $p=0.096$. 
Both $p-$values are clearly larger than that of the 
likelihood ratio test and neither is significant at $\alpha=0.05$. 
This result is consistent with alternative fit indices used in Froh et al paper\citeA{FroFan11}.
